The Weakly Nonlinear Magnetorotational Instability

### 2016-09-02 paper/widegap_mri_paper.tex
make future work an itemized list.

\makeatletter
\setlength{\@fptop}{0pt}
\setlength{\@fpbot}{0pt}
\setlength{\@fpsep}{0pt}
\makeatother

@@

Our investigation is astrophysically motivated, but we also intend our theory to be relevant to laboratory experiments. Several experimental efforts are attempting to observe the MRI in the laboratory, which will allow the study of a crucial astrophysical phenomenon in a controlled setting. Unfortunately, detection of the MRI has so far proven elusive. \citei{Sisan:2004ig} claimed to detect the MRI in a spherical Couette flow, but most likely detected unrelated MHD instabilities instead \citep{Hollerbach:2009ig,Gissinger:2011td}. Most relevant to our work is the Princeton Plasma Physics Laboratory (PPPL) MRI Experiment, a liquid gallium TC flow with an axial magnetic field \citep{Ji:2001kd}. There has been some theoretical work designed to complement the Princeton MRI experiment involving direct numerical simulation of the experimental conditions, much of it focused on the specific challenges in identifying MRI signatures despite apparatus-driven spurious flows (e.g. \citei{Gissinger:2012gc}). The vertical endcaps on a laboratory MRI apparatus drive meridional flows which both inhibit the excitement of MRI and obscure its detection. The Princeton MRI experiment employs split, independently rotating endcaps to mitigate these flows (\citei{Schartman:2009df}). Our work assumes an infinite vertical domain, an idealization that is theoretically expedient but experimentally impractical. Such an approach changes the symmetry properties of the solution significantly, and in the much better studied hydrodynamic case this leads to significant differences even for TC devices with very large aspect ratios \citep{Lopez:2005}. However, our setup is designed to be an accurate treatment of the radial dimension of the flow in a TC apparatus like the one used in the Princeton MRI experiment.

This realistic radial treatment means that we account for the curvature of the flow in a cylindrical apparatus. Many investigations of the MRI use the ``narrow gap" approximation, in which the radial extent of the fluid channel is taken to be much smaller than the radius of curvature. That is, for a center channel radius $r_0$ bounded by inner and outer radii $r_1$ and $r_2$, respectively, the narrow gap approximation applies when $r_0 \gg (r_2 - r_1)$. The narrow gap approximation simplifies the MRI equations by excluding curvature terms, because the flow through a narrow gap can be taken to be approximately linear in $\phi$, i.e. Cartesian. Previous investigations into the weakly nonlinear behavior of the MRI have used this narrow gap approximation (\citei{Umurhan:2007dz}, \citei{Umurhan:2007hs}, \citei{Clark:2016}). In this work we undertake the first (to our knowledge) weakly nonlinear analysis of the MRI in the wide gap regime, where the channel width may be comparable to or larger than its distance from the center of rotation.

@@

The theory presented here may be extended in several ways. 
- detailed comparison to specific experimental apparatus. will gle dynamics be visible? 
- the vertical dimension is incompletely modeled - stratification, etc
- explore saturation property dependence of system on rotation profile 
- different field geometries (e.g. radially varying)
- other radial boundary conditions
- drop axisymmetry
- comparison to simulation in both weakly and strongly nonlinear regimes

>>>>

\makeatletter
\setlength{\@fptop}{0pt}1
\setlength{\@fpbot}{0pt}
\setlength{\@fpsep}{0pt}
\makeatother

@@

Our investigation is astrophysically motivated, but we also intend our theory to be relevant to laboratory experiments. Several experimental efforts are attempting to observe the MRI in the laboratory, which will allow the study of a crucial astrophysical phenomenon in a controlled setting. Unfortunately, detection of the MRI has so far proven elusive. \citei{Sisan:2004ig} claimed to detect the MRI in a spherical Couette flow, but most likely detected unrelated MHD instabilities instead \citep{Hollerbach:2009ig,Gissinger:2011td}. Most relevant to our work is the Princeton Plasma Physics Laboratory (PPPL) MRI Experiment, a liquid gallium TC flow with an axial magnetic field \citep{Ji:2001kd}. There has been some theoretical work designed to complement the Princeton MRI experiment involving direct numerical simulation of the experimental conditions, much of it focused on the specific challenges in identifying MRI signatures despite apparatus-driven spurious flows (e.g. \citei{Gissinger:2012gc}). The vertical endcaps on a laboratory MRI apparatus drive meridional flows which both inhibit the excitement of MRI and obscure its detection. The Princeton MRI experiment employs split, independently rotating endcaps to mitigate these flows (\citei{Schartman:2009df}). Our work assumes an infinite vertical domain, an idealization that is theoretically expedient but experimentally impractical. Such an approach changes the symmetry properties of the solution significantly, and in the much better studied hydrodynamic case this leads to significant differences even for TC devices with very large aspect ratios \citep{Lopez:2005}. Nevertheless, this study represents a first step in understanding the saturation of global, MRI unstable TC flow without the additional complication of vertical endcap effects.

Our radial treatment includes the curvature of the flow in a cylindrical apparatus. Many investigations of the MRI use the ``narrow gap'' approximation (the shearing box is a narrow gap without boundary walls), in which the radial extent of the fluid channel is taken to be much smaller than the radius of curvature. That is, for a center channel radius $r_0$ bounded by inner and outer radii $r_1$ and $r_2$, respectively, the narrow gap approximation applies when $r_0 \gg (r_2 - r_1)$. The narrow gap approximation simplifies the MRI equations by excluding curvature terms, because the flow through a narrow gap can be taken to be approximately linear in $\phi$, i.e. Cartesian. Previous investigations into the weakly nonlinear behavior of the MRI have used this narrow gap approximation (\citei{Umurhan:2007dz}, \citei{Umurhan:2007hs}, \citei{Clark:2016}). In this work we undertake the first (to our knowledge) weakly nonlinear analysis of the MRI in the wide gap regime, where the channel width may be comparable to or larger than its distance from the center of rotation.

@@

There are several avenues for future work, including
\begin{itemize}
\item detailed comparison to specific experimental apparatus. will gle dynamics be visible? 
\item the vertical dimension is incompletely modeled - stratification, etc
\item explore saturation property dependence of system on rotation profile 
\item different field geometries (e.g. radially varying)
\item other radial boundary conditions
\item drop axisymmetry
\item comparison to simulation in both weakly and strongly nonlinear regimes
\end{itemize}
